{\huge Supplements for: Changes and trends in budburst and leaf flush across Europe and North America} \\A meta-analysis of local adaptation in spring phenology studies

### 2024-02-17 Docs/manuscriptRev.Rnw
revised ms

%%% start preambling . . .  %%%
\documentclass{article}
% required

@@

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2023}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
% \doublespacing
\setstretch{1.5} %1.5 spacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\noindent {\small \it
$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
\end{center}
\medskip
\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\

\noindent Additional information: Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\

\noindent Running head: Clines in spring phenology

@@

\begin{abstract} %EWMFeb5 -- edited to define provenance differences 
Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects (i.e. population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
\end{abstract} %Spring tree phenology is critical to forecasting ... Despite a long history of study, however, the balance of local adaptation versus plasticity in determining environmental responses is not well understood, especially for spring events. We conducted the first cross-continental meta-analysis of published studies from the peer-reviewed literature that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data  for  seven angiosperm and eight gymnosperm species from 17 studies, from 384 North American provenances and 101 European provenances with observations from 1962 to 2019.  We found ... This is important because...

\noindent \emph{Keywords: budburst, leafout, budset, senescence, climate change, common gardens, deciduous and evergreen trees, provenance, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract

@@

%EWMFeb5 -- edits below to define provenance; and added a reference to 'local adaptation'
Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees’ phenotypes \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier stopping of growth (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in North America (Fig.\ref{figure:map_gardens} \& Table.S\ref{supplement-table:all_studies} in Supporting Information). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}.

@@

To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May from 2011 to 2020 for all provenances and gardens. We extracted data from E-OBS for European locations and used the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}. Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months for our 10 year period, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.  % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C). % See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.

@@

Overall, spring events were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects. For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
%EWMFeb5 -- more edits here about angio/gymno
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous. As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk. In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation. Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data. %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}. Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.\ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g., \emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly, a critical need for climate change forecasting.

@@

\clearpage
\section{Figures}
%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies. Note: map lines do not necessarily depict accepted national boundaries.}
    \label{figure:map_gardens}
\end{figure}

>>>>

%%% start preambling . . .  %%%
\documentclass[12pt]{article}
% \setmainfont{Times New Roman}
% required

@@

% recommended! Uncomment the below line and change the path for your computer!
\SweaveOpts{prefix.string=/Users/alina/Documents/git/localadaptclim/Docs/figure_ms/demoFig, eps=FALSE} 
%put your figures in one place! Also, note that here 'figures' is the folder and 'demoFig' is what each 
% figure produced will be titled plus its number or label (e.g., demoFig-nqpbetter.pdf')
% make your captioning look better
\usepackage[small]{caption}
\setlength{\captionmargin}{30pt}
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{10pt}
% optional: muck with spacing
\topmargin -1.5cm        
\oddsidemargin 0.5cm   
\evensidemargin 0.5cm  % same as oddsidemargin but for left-hand pages
\textwidth 15.59cm
\textheight 21.94cm 
% \renewcommand{\baselinestretch}{1.5} % 1.5 lines between lines
\parindent 0pt		  % sets leading space for paragraphs
% optional: cute, fancy headers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LO]{2024}
\fancyhead[RO]{Manuscript}
\usepackage{lineno} %alinaSep22: this is to number each line (GCB requires it). However, I dont know how not to number the title page, and I also would like the number to appear bigger. Can lizzie fix this please?
\usepackage{setspace}
% Using \doublespacing in the preamble changes the text to double-line spacing
% \doublespacing
\setstretch{1.5} %1.5 spacing
% more optionals! %
% \usepackage[hyphens]{url} % this wraps my URL versus letting it spill across the page, a bad habit LaTeX has
%%% end preambling. %%%

@@

\noindent {\small \it
$^1$ Forest Resources Management, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4\\
$^2$ Forest \& Conservation Sciences, Faculty of Forestry, University of British Columbia, 2424 Main Mall, Vancouver, BC V6T 1Z4}
\end{center}
\medskip
\noindent{\bf Corresponding author:} Z. A. Zeng, see $^{1}$ above ; E-mail: alinazengziyun@yahoo.com\\
\noindent{\bf Additional information:} Zeng (0009-0000-4647-9457); Wolkovich (0000-0001-7653-893X)\\
\noindent{\bf Running head:} Clines in spring phenology\\
\noindent{\bf Word count:} Summary - 198, Introduction - 1077, Materials and Methods - 821, Results - 1012, Discussion - 1400\\
\noindent{\bf Number of figures:} 4 (Among these, Figure 1 and 2 should be published in color)\\
1. Figure 1: Map showing the distribution of common gardens and provenances.\\
2. Figure 2: Event day of year in relation to provenance latitude and MAT.\\
3. Figure 3: Effects of latitude on spring and fall event day of year depending on continent and species leaf type.\\
4. Figure 4: Effects of MAT on spring and fall event day of year depending on continent and species leaf type.\\
\noindent{\bf Supporting Information (brief legends):}\\
1. Methods S1: Additional methods.\\
2. Table S1: Showing all publications included in meta-analysis.\\
3. Table S2-5: Showing summary of model estimates.\\
4. Figures S1-3: Supporting figures.

@@

% \begin{abstract} %EWMFeb5 -- edited to define provenance differences

\begin{itemize}
  \item Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation vs. through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across previous studies.
  \item We synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American and 101 European provenances (i.e. populations) with observations from 1962 to 2019, alongside fall event data when provided.
  \item Across continents, we find no evidence of provenance effects (i.e. population differences) in spring phenology, but strong clines with latitude and mean annual temperature (MAT) in fall. These effects, however, appear to diverge by continent and species type (gymnosperm vs. angiosperm), with particularly pronounced clines in North America in fall.
  \item Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events.
\end{itemize}
% Forecasting the biological impacts of climate change requires understanding how species respond to warmer temperatures through inter-annual flexible variation versus through adaptation to local conditions. Yet, we often lack this information entirely or find conflicting evidence across studies. The latter is the case for shifts in spring phenology---one of the most reported and consistent impacts of anthropogenic climate change, and also one of the most critical to forecasting, given its role in carbon sequestration. Some common garden studies have found evidence of important provenance effects (i.e. population differences), which suggest there may be local adaptation in the underlying cues of spring phenology and mirrors findings for fall events, while other studies find no evidence. Here, we synthesize common garden studies across Europe and North America that reported spring event dates for a mix of angiosperm and gymnosperm tree species in the northern hemisphere, capturing data from 384 North American provenances and 101 European provenances with observations from 1962 to 2019, alongside fall event data when provided. Across continents, we find no evidence of provenance effects in spring phenology, but strong clines with latitude and mean annual temperature (MAT) for fall events. These effects, however, appear to diverge by continent and species type (gymnosperm versus angiosperm), especially for fall events where clines with latitude and MAT are much stronger in North America. Our results suggest flexible, likely plastic responses, in spring phenology with warming, and potential limits---at least in the short term---due to provenance effects for fall phenology. They also highlight that, after over 250 years of common garden studies on tree phenology, we still lack a holistic predictive model of clines across species and phenological events. 
% \end{abstract}

\noindent \emph{Keywords: budburst, budset, climate change, common gardens, deciduous and evergreen trees, leafout, senescence, spring phenology}\\ %EWMFeb5 -- if you have an abundance of keywords, remove ones that are redundant with title/abstract

@@

%EWMFeb5 -- edits below to define provenance; and added a reference to 'local adaptation'
Studies of adaptation versus plasticity in tree phenology have been conducted for centuries \citep{Cleland:2007or}, through common garden studies. In these studies---conducted often for forestry purposes---researchers grow trees of different geographical origins (called `provenances' often in forestry) under the same environmental conditions to disentangle the effects of environmental and genetic variation on trees' phenotypes \citep{AitkenBemmels16, Alberto13}. Such work has established common clines in fall phenology suggestive of local adaptation, as source locations with shorter growing seasons (poleward and higher elevations) exhibit earlier growth cessation (such as budset). Research has connected these clines to an underlying proximate mechanism of changing photoperiod cues (i.e., shifts in the photoperiod threshold required to trigger budset), driven by adaptation to the local growing season \citep{Alberto13,Savolainen07}. In contrast, spring phenology appears more plastic \citep{AitkenBemmels16} and determined more strongly by temperature \citep{flynn18}. Many studies, however, have argued that spring phenology shows levels of adaptation that may be critical to forecasting and mitigation \citep{vitasse2009,Basler:2012}.

@@

Based on these criteria we found 19 common gardens distributed throughout North America and Europe, with the majority of data concentrated in western North America (Fig.\ref{figure:map_gardens} \& Table.S\ref{supplement-table:all_studies} in Supporting Information). From each common garden study we extracted phenological data on spring events (budburst and leaf flush) in DOY and, when present in the same paper, fall events (bud set, leaf senescence, growth cessation, and leaf abscission) by species and the geographic information of provenances and gardens. We used ImageJ (version 1.53k; \citealp{schneider_rasband_eliceiri_2012}) to extract values from figures whenever necessary. For studies that reported event dates relative to a reference date other than 1 January (e.g. \citealp{Rehfeldt1994}), we converted such dates to DOY using the `lubridate' package in R \citep{Grolemund11}.

@@

To examine climate near spring events more explicitly than MAT allows, we used gridded daily temperature data for March-May from 2011 to 2020 for all provenances and gardens. We extracted data from E-OBS for European locations and used the `daymetr' in R for North American locations \citep{cornes2018,hufkens2018}. Then, using these data and the `overlap' package in R, we estimated how much the daily temperatures overlapped between each provenance location and their corresponding gardens across the three months from 2011 to 2020, which we call `climate overlap.' Dataset containing event dates, geographic information, and climatic information of all provenances are archived in  Knowledge Network for Biocomplexity (KNB) \citep{zeng23}.  % Finally, using the daily temperature data, we also calculated growing degree days (GDD), a commonly used heat accumulation measure to forecast phenological development in plants \citep{miller01}, for each provenance and garden, based on the accumulation of mean daily temperatures ($T_{m}$)from 2011 to 2020 above a baseline of 0$^{\circ}$C, from January 1 until budburst and leaf flush with the following formula: GDD = $\sum$($T_{m}$ - $0^{\circ}$C for $T_{m}$ $\ge$ $0^{\circ}$C; 0 for $T_{m}$ $\le$ $0^{\circ}$C). % See notes on 31 Aug 2023 by Lizzie at https://github.com/lizzieinvancouver/localadaptclim/issues/26 for more methods on GDD.

@@

Overall, spring events such as budburst and leaf flush were not related to provenance latitude or MAT, neither across continents (latitude: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(all_species_springpost_fitC_spring_mat),digits = 2), nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(all_species_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:springfall_latmat}, Table.S\ref{supplement-table:model_spring_lat} \& S\ref{supplement-table:model_spring_mat} in Supporting Information), nor within North America (latitude: \Sexpr{format(round(mean(northamerican_springpost_fitC_spring_lat),digits = 2), nsmall=2)} days/degree [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.05), digits = 1)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(mean(northamerican_springpost_fitC_spring_mat),digits = 1)} days/$^{\circ}$C [\Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(northamerican_springpost_fitC_spring_mat, 0.95), digits = 2)}]) or Europe (latitude: \Sexpr{format(round(mean(european_springpost_fitC_spring_lat),digits = 2),nsmall=2)} days/degree [\Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_lat, 0.95), digits = 2)}]; MAT: \Sexpr{format(round(mean(european_springpost_fitC_spring_mat),digits = 2),nsmall=2)} days/$^{\circ}$C [\Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.05), digits = 2)} - \Sexpr{format(quantile(european_springpost_fitC_spring_mat, 0.95), digits = 2)}]) (Fig.\ref{figure:continental_spptype_lat}A \& \ref{figure:continental_spptype_effect}A). Results were similar using other distance metrics in lieu of latitude (see Fig.S\ref{supplement-figure:lat_distance} for results using the difference between provenance and garden latitude, and the spherical distance between provenance and garden).
% Spring events advanced slightly with provenance latitude in Europe spring events are slightly earlier where provenance MAT is lower (i.e., higher, more northern latitudes), though uncertainty intervals strongly overlapped zero. --- no longer the case

@@

While we expected that coarse metrics, such as latitude and MAT, would generally represent how similar the climates are between the provenances and gardens, we also estimated climate overlap in months much closer to the events to further test how much climate similarity between provenances and gardens predicts provenance effects (i.e. differential responses observed among plant populations from different geographical origins). For spring events, we considered overlap across March to May. However, results were not qualitatively different than using MAT (See Fig.S\ref{supplement-figure:overlap_scatterplot} in Supporting Information). We observed very weak effects of climate overlap on spring events (\Sexpr{format(round(abs(mean(all_species_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(all_species_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent increase in climate overlap), nearly identical across angiosperms (\Sexpr{format(round(abs(mean(deciduous_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(deciduous_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)}]) and gymnosperms (\Sexpr{format(round(abs(mean(evergreens_springpost_fitC_spring_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreens_springpost_fitC_spring_overlap, 0.95)), digits = 2),nsmall=2)}]). Fall events advanced as climate overlap declined, but slightly more strongly for gymnosperms (advancing \Sexpr{format(round(abs(mean(evergreenspost_fitC_fall_overlap)),digits = 2), nsmall=2)} [\Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.05)), digits = 2),nsmall=2)} - \Sexpr{format(round(abs(quantile(evergreenspost_fitC_fall_overlap, 0.95)), digits = 2),nsmall=2)}] days per one percent decline in climate overlap) (Fig.S\ref{supplement-figure:overlap_posterior} in Supporting Information).

@@

%EMWSep30: Please double-check we're explaining trends correctly before you submit; I think we are -- but please check closely.
%EWMFeb5 -- more edits here about angio/gymno
Our results highlight that spring events show much weaker clines than fall events in tree species, but suggest important variation between species types in spring events. Angiosperms tended to budburst earlier in provenances that were warmer and more southerly, while gymnosperms trended in the opposite direction. Such differences could be driven by the varying investment strategies, given that all our studied gymnosperms were evergreen and all angiosperms were deciduous. As evergreen species photosynthesize without leafout they generally leafout much later than deciduous species, after most risk of spring frost \citep{panchen14}, and thus may avoid frost risk. In contrast, deciduous species may tend towards earlier leafout in warmer climates to compete best for access to light and other resources \citep{cat2019}. Testing these hypotheses would require more information on frost risk and forest community assembly from across the provenance locations, but seems an important step towards understanding the drivers of this variation. Without a clear mechanism, extrapolating these results to other species or across Europe and North America may be difficult, especially given biases in the data, and the distinct climatic, geographical, and ecological contexts of Europe and North America. Without a clear and universal mechanism driving the observed phenological responses, the direct applicability of our findings to other contexts must be approached with caution. %EWMFeb5 -- trying to address request by R2 about extrapolating while also keeping ms flow. 
%EMWSep4 -- I like the below (and I think it may have been my idea), but we'd likely be asked to test for it ... and I am not 100% sure it makes sense -- both angiosperms and conifers need to activate their xylem to photosynthesize I believe. 
% Meanwhile, xylem anatomy also plays a part in determining spring timing due to the importance of vessel diameter to embolism risks induced by frost in spring. According to \citet{salk20}, species with narrow xylem such as coniferous gymnosperms and diffuse-porous angiosperms did not exhibit the consistent phenological trends shown in ring-porous species. Therefore, there might be additional differences if we have more species and more information on leafout timing and frost risk approach.

@@

%EMWSep4 -- I like to have one paragraph in a paper (usually) focused on biases and limitations, so the below is that paragraph. I also aimed for it to have more of a message specific to this work beyond 'we need more data' (as we always need more data, you can write that in any paper). 
Taken together, these results could explain some of the existing debate on the strength of spring clinal variation, but also highlight how data biases make disentangling the drivers of variation difficult. Angiosperm species showed a weak trend towards earlier budburst in populations from warmer locations. In contrast, fall event clines in Europe were weak, but their fall event type (leaf senescence) is less likely to show clinal variation. Given European studies were biased towards studying angiosperms (3/4 species) that are often canopy-species (e.g., \emph{Fagus}, \emph{Quercus}) these results together could explain a greater focus on clinal variation in spring events across European studies \citep{Basler:2012,gauzere2020,sog08,deans96,von95}. Most North American gardens and provenances included in this study were limited to the Pacific Northwest region, and thus tended to focus on species from forests where gymnosperms are almost always the canopy species (\emph{Pseudotsuga, Tsuga, Picea, Pinus}) and angiosperms are much smaller, more opportunistic species (\emph{Alnus, Populus}). This might suggest a far different pattern if gardens included more evenly sampled North American tree species (Fig.\ref{figure:map_gardens}). Given the tremendous investment required for such gardens, however, it is not surprising they are often limited to one species---most often those relevant for forestry (e.g., \emph{Pseudotsuga menziesii, Tsuja plicata, Tsuga heterophylla, Picea sitchensis, Pinus albicaulis}). Our results, however, highlight the limited inference we may gain from such an approach to understand trends across species more broadly, a critical need for climate change forecasting.

@@

\clearpage
\section{Figures}
%alinaSep22: i dont know how to force all figures to be in this section instead of in between texts
%EMWSep30 -- you just move them here! Please check I put them in the right order; I also added a \clearpage which can help with formatting looking clean (see if it works, you can remove if not). 
\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/map_gardens.jpg}
    \caption{Distribution of common gardens (denoted as stars) and provenances (denoted as circles) included in this meta-analysis. The distribution was skewed toward North America (12 North American studies versus 5 European studies). See Table.S\ref{supplement-table:all_studies} in Supporting Information for sourcing information on selected studies. Note: map lines do not necessarily depict accepted national boundaries. European sites are made bold and denoted by an asterisk (*). }
    \label{figure:map_gardens}
\end{figure}

\begin{figure}[!h] 
    \centering
 \includegraphics[width=\textwidth]{..//..//localadaptclim/Docs/figure_ms/springfall_latmat.jpg}
    \caption{Event day of year (DOY) in relation to provenance latitude and MAT, coded by symbol for species and color for garden with linear fits from hierarchical Bayesian models. Spring events shown on top and fall events at the bottom. European gardens and species are made bold and denoted by an asterisk (*). } 
    \label{figure:springfall_latmat}
\end{figure}
